---
title: "copernicus"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stars)
library(copernicus)
```

# copernicus

Provides download, archiving and access to [Copernicus](https://marine.copernicus.eu/) marine local datasets.

### Requirements

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [ncdf4](https://CRAN.R-project.org/package=ncdf4)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [readr](https://CRAN.R-project.org/package=readr)
  + [twinkle](https://github.com/BigelowLab/twinkle)
  

### Installation

```
remotes::install_github("BigelowLab/copernicus")
```

## Downloading

You don't need to download data unless you need to download areas outside of `nwa` `[-77.0, -42.5,  36.5,  56.7]` See the [wiki](https://github.com/BigelowLab/copernicus/wiki/Downloading) for hints.

## Local dataset

We maintain a local dataset for the Northwest Atlantic (`nwa` `[-77.0, -42.5,  36.5,  56.7]`).  In particular we have the [GLOBAL_ANALYSIS_FORECAST_PHY_001_024](https://resources.marine.copernicus.eu/?option=com_csw&view=details&product_id=GLOBAL_ANALYSIS_FORECAST_PHY_001_024) for 10 surface water variables and bottom temp going back to January 1, 2019.

### Path

```{r path, message = FALSE}
library(dplyr)
library(stars)
library(copernicus)
dataset <- "global-analysis-forecast-phy-001-024"
path <- copernicus::copernicus_path(dataset, "nwa")
```

### Database

The database is very light, just date variable and depth. Depth recorded this way will allow for intermediate depths and depths ranges ('50' or '50-100').

```{r database, message = FALSE}
db <- copernicus::read_database(path)
db
```

### Look Up Table (lut)

The variable names are a bit cryptic (at least to me) so we provide a look-up table that associates the variable name with other information available from the source.

```{r lut, message = FALSE}
lut <- copernicus::read_lut(name = dataset)
print(lut, n = nrow(lut))
```

### Filenames

Filenames can be made with the database and the path.

```{r filenames, message = FALSE}

small_db <- db %>%
  dplyr::filter(var == 'so' & between(date, as.Date("2019-07-01"), as.Date("2019-07-06")))
filenames <- copernicus::compose_filename(small_db, path)
filenames
```

### Read timeseries of one variable

The [stars](https://r-spatial.github.io/stars/) package provides nice input utilities.  We have selected above just one variable for 6 days. So, it would be nice to have just one attribute with 6 bands.

```{r read_one_var, message = FALSE}
x <- stars::read_stars(filenames, along = 'time') %>%
  stars::st_set_dimensions(which = 'time', value = small_db$date)
plot(x)
```

### Read a day of multiple variables

Let select another subset of multiple variables for just one day (we could do multiple variables for multiple days, but, really, this is just a demo.) In this case we want multiple attributes (variables). [stars](https://r-spatial.github.io/stars/) will do that right out of the box.

```{r read_multi_variable, message = FALSE}
ice_db <- db %>% 
  dplyr::filter(var %in% c('usi', 'vsi', 'siconc', 'sithick') & 
                date == as.Date("2019-12-18"))
y <- ice_db %>%
  copernicus::compose_filename(path) %>%
  stars::read_stars() %>%
  setNames(ice_db$var)
plot(y['siconc'], main = 'siconc', axes = TRUE)
plot(y['sithick'], main = 'sithick', axes = TRUE)
```
