---
title: "copernicus"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides access, download and archiving tools for [Copernicus](https://marine.copernicus.eu/) marine datasets using R language. This package has been developed primarily around ...

+ [GLOBAL_MULTIYEAR_PHY_001_030](https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/description) 31 Dec 1992 to a month or more lag

+ [GLOBAL_ANALYSISFORECAST_PHY_001_024](https://data.marine.copernicus.eu/product/GLOBAL_ANALYSISFORECAST_PHY_001_024/description) 31 Oct 2020 to 9 days from present (forecast)


With no or minimal modification it can work with other products.

### Note

In 2023/2024 Copernicus migrated to a new service model; learn more [here](https://help.marine.copernicus.eu/en/articles/7045314-migrating-to-the-new-global-physical-analysis-and-forecasting-system). This migration introduced the [Copernicus Marine Toolbox](https://help.marine.copernicus.eu/en/collections/4060068-copernicus-marine-toolbox) which provides a Python API and a command line interface (CLI).  This package leverages the latter.


The toolbox is under active development, so if you are having troubles (like we have sometimes!) try re-installing.  We have some notes [here](https://github.com/BigelowLab/copernicus/wiki/Installation-of-%60copernicusmarine%60).

# Copernicus resources

Copernicus serves **so many** data resources; finding what you want can be a challenge.  Check out the new [Marine Data Store](https://marine.copernicus.eu/news/introducing-new-copernicus-marine-data-store). And checkout the listing [here](https://marine.copernicus.eu/about/producers).

## Data catalogs

Like data offerings from [OBPG](https://oceancolor.gsfc.nasa.gov/), Copernicus strives to provided consistent dataset identifiers that are easily decoded programmatically (and with practice by eye). In order to download programmatically you must have the datasetID in hand. Learn more about Copernicus [nomenclature rules here](https://help.marine.copernicus.eu/en/articles/6820094-how-is-defined-the-nomenclature-of-copernicus-marine-data#h_34a5a6f21d).  We leverage the [data catalogs] which link a **product** to one or more of its **datasets**.  Did you catch that?  Copernicus distributes **products** each of which comes as one or more **datasets**.  There's mor eon that later.

## `get` or `subset`

The [Copernicus Marine Toolbox](https://help.marine.copernicus.eu/en/collections/4060068-copernicus-marine-toolbox) command-line application, `copernicus-marine` provides two primary methods for donwloading data: `get` and `subset`. `get` is not well documented, but subset does what it implies - subsetting resources by variable, spatial bounding box, depth and time.  This package only supports `subset`.

## Requirements

  + [R v4.1+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [ncdf4](https://CRAN.R-project.org/package=ncdf4)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [readr](https://CRAN.R-project.org/package=readr)
  

## Installation

```
remotes::install_github("BigelowLab/copernicus")
```

## Configuration

You can preconfigure a credentials file (required) and a path definition file (optional) to streamline accessing and storing data.

### Configure credentials

You must have credentials to access Copernicus holdings - if you don't have them now please request access [here](https://data.marine.copernicus.eu/register).  Go straight to it [here](https://help.marine.copernicus.eu/en/articles/8185007-copernicus-marine-toolbox-credentials-configuration).

### Configure data path

If you plan to use our directory-driven database storage system then you should set the root path for the data directory; keep in mind that you can always change or override it. We don't actually run this in the README, but you can copy-and-paste to use in R.  Again, replace the path with one suiting your own situation.

```
copernicus::set_root_path("/the/path/to/copernicus/data")
```

Once it is set, you shouldn't have to set it again (unless you want to change the path).

#### Configure the application path

If you are using R within a RStudio session, you may encounter issues where `system()` and `system2()` can't' find the `copernicusmarine` application. This is not the case when you run R outside of the RStudio context. Technically, this is a environmental path issue, which you can remedy by providing the full path specification for the `app` argument to the function `build_cli_subset()`.  By default, `app = 'copernicusmarine`, but you may need to include the full path specification.  We provide a mechanism for storing this path in a configuration file once, and then it will work without issue in subsequent sessions of R. Here's how we set ours. 

First determine the app path in the terminal session (outside of RStudio context).

```
$ which copernicusmarine
/opt/copernicus/bin/copernicusmarine
```

Then set the path.
```{r}
copernicus::set_copernicus_app("/opt/copernicus/bin/copernicusmarine")
```

This is optional (but worth it if you operate within RStudio). You can retrieve the application path with `get_copernicus_app()`, which defaults to `copernicusmarine` if you didn't set the path.

```{r}
copernicus::get_copernicus_app()
```

## Product catalog

You can download a product catalog for local storage.

```{r}
suppressPackageStartupMessages({
  library(copernicus)
  library(stars)
})

ok = copernicus::fetch_product_catalog(product_id = "GLOBAL_ANALYSISFORECAST_BGC_001_028")
```

This downloads into a "catalogs" directory within your data directory Now read it in.

```{r read_catalog}
x = copernicus::read_product_catalog(product_id = "GLOBAL_ANALYSISFORECAST_BGC_001_028")
x
```

By default this provides a flattened table of available datasets along with tables of variables for each (if any).  Here's the first dataset (a constituent of product suite)

```{r variable_table}
dplyr::filter(x, dataset_id %in% "cmems_mod_glo_bgc-bio_anfc_0.25deg_P1D-m") |>
  dplyr::glimpse()
```


## Which products?  Which datasets?

[CMEMS](https://data.marine.copernicus.eu/products) offers many products - which ones have we set up for this package? (Note, it's subject to change).

[GLOBAL_MULTIYEAR_PHY_001_030](https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/description) 31 Dec 1992 to a month or more lag

[GLOBAL_ANALYSISFORECAST_PHY_001_024](https://data.marine.copernicus.eu/product/GLOBAL_ANALYSISFORECAST_PHY_001_024/description) 31 Oct 2020 to 9 days from present (forecast)

[GLOBAL_ANALYSISFORECAST_BGC_001_028](https://data.marine.copernicus.eu/product/GLOBAL_ANALYSISFORECAST_BGC_001_028/description) 2021-10-01 to 9 days from present


## Fetching data

To fetch data we'll focus on [ocean physics daily forecast](https://data.marine.copernicus.eu/product/GLOBAL_ANALYSISFORECAST_PHY_001_024/download?dataset=cmems_mod_glo_phy-cur_anfc_0.083deg_P1D-m) which serves daily mean sea surface currents. We'll define a date range and the bounding box that covers the Gulf of Maine (gom), and we'll confine the request to just the surface data.

```{r fetch}
suppressPackageStartupMessages({
  library(stars)
  library(copernicus)
  library(dplyr)
})
product_id = "GLOBAL_ANALYSISFORECAST_PHY_001_024"
dataset_id = "cmems_mod_glo_phy-cur_anfc_0.083deg_P1D-m"    
vars = c("uo","vo")
bb = c(xmin = -72, xmax = -63, ymin = 39, ymax = 46)
path = copernicus_path(product_id, "gom") |>
  make_path()
depth = c(0,1) # just the top 1 meter
time = c(0, 9) + Sys.Date()  # today - and a little ahead window
ofile = copernicus_path("tmp", 
                        paste0(product_id, "__", dataset_id, ".nc"))
ok = download_copernicus_cli_subset(dataset_id = dataset_id, 
                                   vars = vars, 
                                   depth = depth,
                                   bb = bb, 
                                   time = time, 
                                   ofile = ofile)
x = stars::read_stars(ofile)
x
```



```{r first_plot}
plot(x['uo'], axes = TRUE)
```
